---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data
```{r}
f<-unzip("activity.zip")
data<-read.csv(f)
rm(f)
data$dates<-strptime(data$date, "%Y-%m-%d")
uniqueDates <- unique(data$dates)
uniqueIntervals <- unique(data$interval)
```


## What is mean total number of steps taken per day?
### Data preparation
```{r}
stepsSpl<-split(data$steps, data$dates$yday)
sumStepsPerDay<-sapply(stepsSpl, sum, na.rm=TRUE)
```

### Total number of steps taken each day
```{r}
plot(uniqueDates, sumStepsPerDay, main="Histogram of steps per day", xlab="Date", ylab="Frequency", type="h", lwd=4, col="blue")
```

### Mean and median steps per day
```{r}
meanStepsPerDay <- sapply(stepsSpl, mean, na.rm=TRUE)
meanDataFrame <- data.frame(date=uniqueDates, meanStepsPerDay=meanStepsPerDay, row.names=NULL)
medianStepsPerDay <- sapply(stepsSpl, median, na.rm=TRUE)
medianDataFrame <- data.frame(date=uniqueDates, medianStepsPerDay=medianStepsPerDay, row.names=NULL)
```

#### Mean steps per day
```{r}
print(meanDataFrame)
```

#### Median steps per day
```{r}
print(medianDataFrame)
```

## What is the average daily activity pattern?
```{r}
intervalSplit <- split(data$steps, data$interval)
averageStepsPerInterval <- sapply(intervalSplit, mean, na.rm=TRUE)
plot(uniqueIntervals, averageStepsPerInterval, type="l",main="Average number of steps per interval.\n (blue line indicates the interval with maximum steps on average)", xlab="Interval", ylab="Average # of steps ", lwd=2, col="red")
maxIntervalDays <- max(averageStepsPerInterval, na.rm=TRUE)
maxIndex <- as.numeric(which(averageStepsPerInterval == maxIntervalDays))
maxInterval <- uniqueIntervals[maxIndex]
abline(v=maxInterval, col="blue", lwd=3)
```

## Imputing missing values
### Amount of missing values
```{r}
isna<- is.na(data$steps)
print(sum(isna))
```

### Dealing with missing values - replace NAs with means
```{r}
# Initially replace NaNs in mean steps per day with 0s
meanStepsPerDay[is.nan(meanStepsPerDay)]<-0
# To deal with NaNs we will replace them with means. To do that
# - create a vector with mean values for each day
meanValues <- rep(meanStepsPerDay, length(data$steps)/length(meanStepsPerDay))
# - Create new dataset.
dataNew<-data
# Now replace these values with their corresponding mean
dataNew$steps[is.na(data$steps)] <- meanValues[is.na(data$steps)]
```

### Histograms to compare of the total number of steps taken each day before and after missing values are imputed
```{r}
if(!require(ggplot2))
  stop("you need to install ggplot2")
if(!require(ggplot2))
  stop("you need to install gridextra")
stepsSplNew<-split(dataNew$steps, dataNew$dates$yday)
sumStepsPerDayNew<-sapply(stepsSplNew, sum, na.rm=TRUE)
library(ggplot2)
p<-ggplot(data=data.frame(uniqueDates, sumStepsPerDay), aes(x=uniqueDates, y=sumStepsPerDay))+geom_bar(stat="identity")+labs(y = "Steps per day", x="Dates", title="Steps per day\nincluding unknown values")
p2<-ggplot(data=data.frame(uniqueDates, sumStepsPerDayNew), aes(x=uniqueDates, y=sumStepsPerDayNew))+geom_bar(stat="identity")+labs(y = "Steps per day", x="Dates", title="Steps per day;\nunknown values replaced\nwith mean")
library(gridExtra)
grid.arrange(p, p2, ncol=2)
```

### Mean and median steps per day with missing values imputed
```{r}
meanStepsPerDayNew <- sapply(stepsSplNew, mean, na.rm=TRUE)
meanDataFrameNew <- data.frame(date=uniqueDates, meanStepsPerDay=meanStepsPerDayNew, row.names=NULL)
medianStepsPerDayNew <- sapply(stepsSplNew, median, na.rm=TRUE)
medianDataFrameNew <- data.frame(date=uniqueDates, medianStepsPerDay=medianStepsPerDayNew, row.names=NULL)
```

#### Mean steps per day (with NaNs replaced with means)
```{r}
print(meanDataFrameNew)
```

#### Median steps per day (with NaNs replaced with means)
```{r}
print(medianDataFrameNew)
```

## Are there differences in activity patterns between weekdays and weekends?
### Prepare datasets to compare
```{r}
clwdays<-rep(2, length(data$dates))
clwdays[data$dates$wday>=1 & data$dates$wday<=5]<-1
dataNew$dateType<-factor(clwdays, levels=c(1,2), labels=c("Weekdays", "Weekends"))
dataNewWeekdays <- dataNew[dataNew$dateType == "Weekdays", ]
dataNewWeekends <- dataNew[dataNew$dateType == "Weekends", ]

dataNewSplitWeekdays <- split(dataNewWeekdays$steps, dataNewWeekdays$interval)
dataNewSplitWeekends <- split(dataNewWeekends$steps, dataNewWeekends$interval)

# Find the average for each interval
meanStepsPerWeekdayInterval <- sapply(dataNewSplitWeekdays, mean)
meanStepsPerWeekendInterval <- sapply(dataNewSplitWeekends, mean)
```

### Average number of steps taken per 5-minute interval across weekdays and weekends
```{r}
par(mfcol=c(2,1))
plot(uniqueIntervals, meanStepsPerWeekdayInterval, type="l",
     main="Average number of steps - weekdays", 
     xlab="Interval", ylab="Average # of steps", 
     lwd=2, col="blue")
plot(uniqueIntervals, meanStepsPerWeekendInterval, type="l",
     main="Average number of steps - weekends", 
     xlab="Interval", ylab="Average # of steps", 
     lwd=2, col="red")
```
